---
title: "Comparison of WHO and The Economist Excess Mortality impact on vaccine deaths averted"
author:
  - Greg Barnsley:
      email: g.barnsley@imperial.ac.uk
      institute: [ICL]
      correspondence: true
institute:
  - ICL: Imperial College London
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    word_document
bibliography: references.bib
params:
  gather_raw_data: FALSE
  parallel_: TRUE
  generate_model_fits: TRUE
---

<!-- The following code chunk defines some general settings how code chunks should behave. -->

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/",
  dpi = 300
)

library(tidyverse)
library(lubridate)
library(readxl)
library(keras)
```

<!-- The actual document text starts here: -->

# Introduction

<!-- do last then update description and upload to github -->

# Comparison of Estimates

```{r load Economist}
econ_raw_file_path <- file.path(here::here(), "analysis/data/raw_data/economist_raw.csv")
if(params$gather_raw_data){
  #save to the raw_data_folder
  download.file("https://raw.githubusercontent.com/TheEconomist/covid-19-the-economist-global-excess-deaths-model/5fa640a2cc53378b10070e2be505e80d8448ee39/output-data/export_country.csv", econ_raw_file_path)
}
#load raw version from data_raw
econ_df <- read_csv(econ_raw_file_path)
#process to get what we need, date + iso3c + combined median & true value
econ_df <- econ_df %>% 
  mutate(economist_estimated = is.na(daily_excess_deaths)) %>% 
  select(date, iso3c, estimated_daily_excess_deaths, economist_estimated) %>% 
  rename(economist_excess_deaths = estimated_daily_excess_deaths)
```

```{r load WHO}
who_raw_file_path <- file.path(here::here(), "analysis/data/raw_data/who_raw.zip")
if(params$gather_raw_data){
  #save to the raw_data_folder
  download.file("https://cdn.who.int/media/docs/default-source/world-health-data-platform/covid-19-excessmortality/2022-03-25_covid-19_gem.zip", who_raw_file_path)
}
#extract files
unzip(who_raw_file_path, "WHO_COVID_Excess_Deaths_EstimatesByCountry.xlsx", exdir = here::here("temp"))
#load raw version from unzipped
who_df <- read_xlsx(here::here("temp/WHO_COVID_Excess_Deaths_EstimatesByCountry.xlsx"),
                    sheet = "Country by year and month", skip = 12)
#remove temp folder
unlink(here::here("temp"), force = TRUE, recursive = TRUE)
#process to get what we need, date + iso3c + excess
who_df <- who_df %>% 
  mutate(who_estimated = type == "Predicted",
         date = as.Date.character(paste0(as.character(year), "-", as.character(month), "-01"))) %>% 
  select(date, iso3, excess.mean, who_estimated) %>% 
  rename(iso3c = iso3, who_excess_deaths = excess.mean) %>% 
  #Scale the December value so that its representative up to 2021-12-08
  mutate(
    who_excess_deaths = if_else(
      date == "2021-12-01",
      who_excess_deaths*(8/31),
      who_excess_deaths
    )
  )
```

```{r combine WHO ECON}
#they are in different formats atm so we will convert economist into monthly
econ_df <- econ_df %>% 
  group_by(iso3c) %>% 
  arrange(iso3c, date) %>% 
  #assume daily econo estimate date is the middle date for the week
  #first get the first day of the week for that date
  mutate(week_start_date = floor_date(date, "week"),
         #only update this there's a whole week between, this way we still apply the final estimate
         date = if_else(diff(c(min(week_start_date) - 7, week_start_date)) == 7,
                        week_start_date, date)) %>% 
  complete(date = seq(min(date), max(week_start_date) + 7, 1)) %>% 
  fill(economist_excess_deaths, economist_estimated, .direction = "down") %>% 
  #remove values that fall outside our fitting period
  filter(date > "2019-12-31" & date <= "2021-12-08") %>% 
  #now group into months
  mutate(month = month(date),
         year = year(date)) %>% 
  group_by(iso3c, year, month) %>% 
  summarise(economist_excess_deaths = sum(economist_excess_deaths),
            economist_estimated = any(economist_estimated),
            date = min(date), .groups = "drop") %>% 
  select(!c(month, year))
#now only keep countries in both
in_both <- intersect(intersect(econ_df$iso3c, who_df$iso3c), squire::population$iso3c)
combined_df <- econ_df %>% 
  filter(iso3c %in% in_both) %>% 
  full_join(
    who_df %>% 
      filter(iso3c %in% in_both)
  ) %>% #add pops %>% 
  filter(!iso3c %in% c("CHN")) %>% 
  left_join(
    squire::population %>% 
      group_by(iso3c) %>% 
      summarise(
        pop = sum(n)
      )
  )
#save to data folder
write_csv(combined_df %>% 
            rename(month = date),
          file.path(here::here("analysis/data/derived_data/combined_estimates.csv")))
```

```{r totals numeric}
totals_overall <- combined_df %>% 
  summarise(
    across(
      c(economist_excess_deaths, who_excess_deaths),
      ~format(signif(sum(.x), 4), big.mark = ",")
    )
  ) %>% 
  as.list()

totals_2021 <- combined_df %>% 
  filter(date >= "2020-12-01") %>% 
  summarise(
    across(
      c(economist_excess_deaths, who_excess_deaths),
      ~format(signif(sum(.x), 4), big.mark = ",")
    )
  ) %>% 
  as.list()
```

```{r totals plot}
#only interested in values where one or the other are estimated
est_total_df <- combined_df %>% 
  filter(economist_estimated | who_estimated) %>% 
  group_by(iso3c) %>% 
  summarise(
    across(c(economist_excess_deaths, who_excess_deaths),
           ~sum(.x)),
    pop = pop[1]
  )

total_comp_df <- est_total_df %>% 
  mutate(
           economist_excess_deaths = economist_excess_deaths/pop * 1000,
           who_excess_deaths = who_excess_deaths/pop * 1000,
           label = if_else(
              who_excess_deaths > economist_excess_deaths + 2.5 |
                who_excess_deaths < economist_excess_deaths -3,
              countrycode::countrycode(iso3c, origin = "iso3c", destination = "country.name"),
              ""
            )
           )
total_comp_plot <- ggplot(total_comp_df) +
  geom_abline(intercept = 0, slope = 1, alpha = 0.5, linetype = "dashed") + 
  geom_smooth(aes(x = economist_excess_deaths, y = who_excess_deaths), method = "lm", fullrange = TRUE, level = 0, colour = "black") + 
  geom_point(aes(x = economist_excess_deaths, y = who_excess_deaths, colour = squire.page::get_WHO_region(iso3c)), alpha = 0.9) + 
  geom_text(aes(x = economist_excess_deaths, y = who_excess_deaths, label = label), nudge_y = -0.25) +
  ggpubr::theme_pubr() + 
  labs(x = "Economist Estimated Excess Mortality per 1000",
       y = "WHO Estimates Excess Mortality per 1000",
       colour = "WHO Region:")

new_gradient <- lm(who_excess_deaths ~ economist_excess_deaths, total_comp_df %>% 
  filter(!iso3c %in% c("DJI", "IRQ", "SDN")))$coefficients[2]

#save to figures
ggsave(here::here("analysis/figures/01_total-comparison.pdf"), total_comp_plot)
```

```{r monthly plot}
monthly_comp_plot <- combined_df %>% 
  group_by(date) %>% 
  summarise(
    across(
      c(economist_excess_deaths, who_excess_deaths),
      ~sum(.x)
    )
  ) %>% 
  mutate(
    ratio = economist_excess_deaths - who_excess_deaths
  ) %>% 
  ggplot() +
  geom_step(aes(x = date, y = ratio)) + 
  ggpubr::theme_pubr() + 
  labs(x = "Date", y = "Monthly Difference between Economist Estimate\nand WHO Estimate") + 
  scale_y_continuous(labels = scales::comma)

ggsave(here::here("analysis/figures/02_monthly-comparison.pdf"), monthly_comp_plot)
```

To briefly compare the WHO estimates to the estimates from The Economist (used 
in the vaccine impact paper) we first align the format of the two datasets.
Given the weekly basis for the Economists' estimates we aggregate these into a
monthly format. Comparing across the average estimates all countries and 
territories that are in both datasets and included in the vaccine impact paper 
(i.e. excluding China and countries with populations less than 90,000), the 
WHO gives `r totals_overall$who_excess_deaths` compared to
`r totals_overall$economist_excess_deaths` from The Economist. Limiting this
to the year 2021 gives us `r totals_2021$who_excess_deaths` compared to
`r totals_2021$economist_excess_deaths`.

```{r, fig.cap = "Comparison of Excess Mortality Estimates per country per 1,000 people. The grey dashed line repesents the gradient of equality and the black line the least squares estimate of the gradient."}
print(total_comp_plot)
```


Comparing this across countries in figure 1, there is a strong positive 
correlation between the two sets of estimates, as to be expected. However, 
there is a trend for The Economist to estimate higher mortality than the WHO,
evidenced by comparing the line of best fit to the slope of equality. There are
several outliers in the this comparison, notably Djibouti, Iraq, and Sudan
where The Economist model estimates very high mortality, higher than any other 
country included (per capita). Iraq, and Sudan were also outliers in the 
vaccine-impact paper as the model could recreate the death curves of neither. 
Excluding these from the line of best fit estimation, gives a gradient of 
`r new_gradient`.

```{r, fig.cap = "Comparison of Global Monthly Excess Mortality between the two estimates."}
print(monthly_comp_plot)
```

Figure 2, taken over all excess mortality (real or estimated) shows a temporal
aspect to the difference between the estimates. With roughly two distinct peaks 
(meaning that the Economist is estimating higher than the WHO) one over the first
half of 2020 and another in the last half of 2021.

# Impact on Deaths Averted

```{r get deaths averted}
if(params$gather_raw_data){
  #calculate deaths averted from the vaccine impact repo
  #we are processing this data because otherwise it would be many largish files
  map_dfr(unique(combined_df$iso3c), function(iso3c){
    dir.create(here::here("temp"))
    exists <- tryCatch(
    walk(c("Baseline_", "No%20Vaccines_"), function(file){
      download.file(
        paste0("https://github.com/mrc-ide/covid-vaccine-impact-orderly/blob/main/data/excess_mortality/counterfactual_data/", 
           file, iso3c, ".Rds?raw=true"), here::here(paste0("temp/",file, ".Rds")),
        mode = "wb"
      )
    }),
    error = function(e){NULL})
    if(!is.null(exists)){
      base <- readRDS(here::here("temp/Baseline_.Rds"))
    cf <- readRDS(here::here("temp/No%20Vaccines_.Rds"))
    unlink(here::here("temp"), force = TRUE, recursive = TRUE)
    tibble(
      date = base$date,
      replicate = base$replicate,
      deaths_averted = cf$deaths - base$deaths,
      deaths = base$deaths,
      iso3c = base$iso3c,
      year = year(base$date),
      month = month(base$date)
    ) %>%
      group_by(replicate, year, month) %>% 
      summarise(
        deaths_averted = sum(deaths_averted),
        deaths = sum(deaths),
        iso3c = iso3c[1],
        date = min(date),
        .groups = "drop"
        ) %>% 
      select(iso3c, replicate, date, deaths, deaths_averted)
    } else {
      NULL
    }
  }) %>% 
    saveRDS(here::here("analysis/data/raw_data/excess_mortality.Rds"))
  #now get the reported deaths data
  map_dfr(unique(combined_df$iso3c), function(iso3c){
    dir.create(here::here("temp"))
    exists <- tryCatch(
    walk(c("Baseline_", "No%20Vaccines_"), function(file){
      download.file(
        paste0("https://github.com/mrc-ide/covid-vaccine-impact-orderly/blob/main/data/reported_deaths/counterfactual_data/", 
           file, iso3c, ".Rds?raw=true"), here::here(paste0("temp/",file, ".Rds")),
        mode = "wb"
      )
    }),
    error = function(e){NULL})
    if(!is.null(exists)){
    base <- readRDS(here::here("temp/Baseline_.Rds"))
    cf <- readRDS(here::here("temp/No%20Vaccines_.Rds"))
    unlink(here::here("temp"), force = TRUE, recursive = TRUE)
    tibble(
      date = base$date,
      replicate = base$replicate,
      deaths_averted = cf$deaths - base$deaths,
      deaths = base$deaths,
      iso3c = base$iso3c,
      year = year(base$date),
      month = month(base$date)
    ) %>%
      group_by(replicate, year, month) %>% 
      summarise(
        deaths_averted = sum(deaths_averted),
        deaths = sum(deaths),
        iso3c = iso3c[1],
        date = min(date),
        .groups = "drop"
        ) %>% 
      select(iso3c, replicate, date, deaths, deaths_averted)
    } else {
      NULL
    }
  }) %>% 
    saveRDS(here::here("analysis/data/raw_data/reported_deaths.Rds"))
}
#load data
deaths_averted <- readRDS(here::here("analysis/data/raw_data/excess_mortality.Rds")) %>% 
  mutate(type = "excess") %>% 
  rbind(
    readRDS(here::here("analysis/data/raw_data/reported_deaths.Rds")) %>% 
      mutate(type = "reported")
  ) %>% 
  filter(!is.na(date)) %>% 
  filter(date > "2019-12-31") %>%
  mutate(date = floor_date(date, "month")) %>%  #ensure this is in terms of whole months
  #drop ones with no vaccinations (i.e. no deaths averted)
  group_by(iso3c) %>% 
  filter(sum(deaths_averted) != 0) %>% 
  group_by(iso3c, date, type)
```

```{r get vaccine data}
#Raw data unavailable from Unicef dashboard https://www.unicef.org/supply/covid-19-vaccine-market-dashboard
simplify_vaccines <- function(x){
  case_when(
      x %in% c("Pfizer/BioNTech", "Moderna") ~ "mRNA",
      x %in% c("Abdala", "Novavax", "ZF2001", "Soberana02", "Soberana Plus", "Razi Cov Pars", "SpikoGen", "EpiVacCorona", "Medigen", "Abdala (Subunit like novavax)", "Anhui ZL (Recombinant Platform)") ~ "Subunit",
      x %in% c("Oxford/AstraZeneca", "Sputnik V", "Sputnik Light", "CanSino", "Covishield") ~ "Adenovirus",
      x %in% c("Sinopharm/Wuhan", "Sinopharm/Beijing", "Sinovac", "Covaxin", "COVIran Barekat", "FAKHRAVAC", "QazVac", "Turkovac", "KoviVac/Chumakov") ~ "Whole Virus",
      TRUE ~ x,
    )
}
dominant_vaccines <- readRDS(here::here("analysis/data/raw_data/dominant_vaccines.Rds"))%>% 
  transmute(
    iso3c = countrycode::countrycode(country, "country.name", "iso3c"),
    #reduce to types
    dominant_vaccine = simplify_vaccines(dominant)
  )
if(params$gather_raw_data){
  url <- xml2::read_html("https://github.com/owid/covid-19-data/tree/master/public/data/vaccinations/country_data")
  ctries <- rvest::html_text(rvest::html_nodes(url, ".js-navigation-open.Link--primary"))
  links <- paste0("https://raw.githubusercontent.com/owid/covid-19-data/master/public/data/vaccinations/country_data/", ctries)
  links <- urltools::url_encode(links)
  dat <- lapply(links, read.csv)
  present_vaccines <- data.frame("country" = gsub(".csv","", ctries, fixed = TRUE),
                 "vaccines" = unlist(lapply(dat, function(x){tail(x$vaccine,1)}))) %>%
    mutate(
      country = case_when(
        country %in% c("England", "Northern Ireland", "Scotland", "Wales") ~ "United Kingdom",
        country == "Timor" ~ "Timor Leste",
        TRUE ~ country
      ),
      country = if_else(
        country == "Kosovo",
        "XKX",
        countrycode::countrycode(country, origin = "country.name", destination = "iso3c")
      )
    ) %>%
    filter(country %in% dominant_vaccines$iso3c) %>%
    rename(iso3c = country) %>%
    unique() %>%
    group_by(iso3c) %>%
    summarise(
      vaccines = paste0(vaccines, collapse = ", ")
    )
  saveRDS(present_vaccines, here::here("analysis/data/raw_data/owid_vaccine_usage.Rds"))
  rm(url, ctries, links, dat)
} else {
  present_vaccines <- readRDS(here::here("analysis/data/raw_data/owid_vaccine_usage.Rds"))
}
#add missing values
present_vaccines <- present_vaccines %>%
  rbind(
    tibble(
      iso3c = c("GUF", "GLP", "MTQ", "MYT", "REU"),
      vaccines = present_vaccines %>%
        filter(iso3c == "FRA") %>%
        pull(vaccines)
    )
  ) %>%
  rbind(
    vaccines = present_vaccines %>%
      filter(iso3c == "GBR") %>%
      mutate(iso3c = "CHI")
  ) %>%
  rbind(
    tibble(
      iso3c = c("GUM", "FSM", "PRI", "VIR"),
      vaccines = c("Pfizer/BioNTech, Moderna, Johnson&Johnson",
                   "Moderna, Johnson&Johnson",
                   "Pfizer/BioNTech, Moderna, Johnson&Johnson",
                   "Pfizer/BioNTech, Moderna")
    )
  )

#split into distinct categories
vaccine_types <- unique(unlist(str_split(unique(present_vaccines$vaccines), ", ")))
#simplify to types
vaccine_cat <- simplify_vaccines(vaccine_types)
unique_cat <- unique(vaccine_cat)
names(unique_cat) <- unique_cat
#reduce
present_vaccines <- present_vaccines %>%
  select(!vaccines) %>%
  cbind(
    map_dfc(unique_cat, function(category){
      reduce(vaccine_types[vaccine_cat == category],
             function(curr, new_vaccine_type){
               curr + str_detect(present_vaccines$vaccines, new_vaccine_type) > 0
             },
             .init = FALSE
      )
    })
  )
rm(vaccine_types, vaccine_cat, unique_cat)

#Also load when variant timings start, no source available generated when model
#fitted to covariants data
variant_timings <- readRDS(here::here("analysis/data/raw_data/variant_characteristics.rds")) %>% 
  map_dfr(function(country) map_dfc(country, ~.x$start_date), .id = "iso3c") %>% 
  transmute(
    iso3c = iso3c,
    across(!iso3c,
           ~as.numeric(.x),
           .names = "start_date_{col}")
  )

#start_date of epidemic (rough for now)
epidemic_start_dates <- deaths_averted %>% 
  group_by(iso3c, replicate, type) %>% 
  filter(deaths > 0) %>% 
  summarise(
    start_date_epidemic = as.numeric(min(date)),
    .groups = "drop"
  )

#well assume this is the first non zero deaths averted in any replicate/type of fit
vaccination_start_dates <- deaths_averted %>% 
  group_by(iso3c) %>% 
  filter(deaths_averted > 0) %>% 
  summarise(start_date_vaccination = min(date))

#again no source, based on formattted OWID data
vaccinations_total <- readRDS(here::here("analysis/data/raw_data/vacc_inputs.rds")) %>% 
  map_dfr(~tibble(first_doses = sum(.x$max_vaccine), second_doses = tail(.x$dose_ratio, 1)*sum(.x$max_vaccine)), .id = "iso3c")

age_baseline_ifr <- nimue:::default_probs()$prob_hosp * 
        (nimue:::default_probs()$prob_severe * nimue:::default_probs()$prob_severe_death_treatment + 
        (1-nimue:::default_probs()$prob_severe) * nimue:::default_probs()$prob_non_severe_death_treatment)
##used to calculate the baseline IFR for the demographic
```

```{r setup deaths array}
#multi dimensional array for deaths by vaccination period and variant period
deaths_array <- deaths_averted %>%
  ungroup() %>%
  #create a unique id for each country, type and replicate
  mutate(
    id = paste(iso3c, type, replicate, sep = "_")
  ) %>%
  left_join(variant_timings, by = "iso3c") %>%
  left_join(vaccination_start_dates, by = "iso3c") %>%
  left_join(
    squire::population %>%
      group_by(iso3c) %>%
      summarise(pop = sum(n)),
    by = "iso3c"
  ) %>%
  mutate(variant = case_when(
    date < start_date_Delta ~ "Wild",
    TRUE ~ "Delta"
  ),
  vaccinations = if_else(
    date < start_date_vaccination,
    "NoVaccine",
    "Vaccine"
  )) %>%
  group_by(id, variant, vaccinations) %>%
  summarise(
    deaths_per_capita = sum(deaths)/pop,
    .groups = "keep"
  ) %>% #complete vaccinations and deaths
  group_by(id) %>%
  complete(vaccinations = c("Vaccine", "NoVaccine"), variant = c("Wild", "Delta"), fill = list(deaths_per_capita = 0)) %>%
  arrange(id, vaccinations, variant)

#standardise by unique combination
death_means <- deaths_array %>%
  group_by(variant, vaccinations) %>%
  summarise(
    mean = mean(deaths_per_capita),
    var = var(deaths_per_capita),
    .groups = "drop"
  ) %>%
  left_join(death_means, by = c("vaccinations", "variant")) %>%
  mutate(
    deaths_per_capita_standardised = (deaths_per_capita - mean)/sqrt(var)
  ) %>%
  select(!c(mean, var, deaths_per_capita)) %>% 
  convert_df_to_array()
```

```{r setup other fitting parameters}
#dataframe for the rest, i.e. parameter values
parameter_data <- deaths_averted %>%
  ungroup() %>%
  mutate(id = paste(iso3c, type, replicate, sep = "_")) %>%
  select(id, iso3c, replicate, type) %>%
  group_by(iso3c) %>%
  #IFR at t0
  mutate(
    baseline_ifr =
      weighted.mean(
        age_baseline_ifr,
        squire::population[squire::population$iso3c == iso3c[1],"n"]
      )
  ) %>%
  left_join(
    squire::population %>%
      group_by(iso3c) %>%
      summarise(pop = sum(n)),
    by = "iso3c"
  ) %>%
  #hopsital and ICU capacity
  mutate(
    hosp_beds = squire::get_healthcare_capacity(
      squire::population$country[squire::population$iso3c == iso3c[1]][1]
    )$hosp_beds,
    ICU_beds = squire::get_healthcare_capacity(
      squire::population$country[squire::population$iso3c == iso3c[1]][1]
    )$ICU_beds
  ) %>%
  ungroup() %>%
  #mixing matrix
  #add vaccine efficacy, for each replicate for each variant
  #epidemic start date
  left_join(epidemic_start_dates, by = c("iso3c", "replicate", "type")) %>%
  #vaccination counts
  left_join(vaccinations_total, by = "iso3c") %>%
  #make per capita
  mutate(across(
    c(first_doses, second_doses),
    ~.x/pop, .names = "{.col}_per_capita")
  ) %>%
  ungroup() %>%
  select(!c(iso3c, replicate, type, pop, first_doses, second_doses))
#standardise some variables
vars_to_standardise <- c("baseline_ifr", "hosp_beds", "ICU_beds", "start_date_epidemic",
                         "first_doses_per_capita", "second_doses_per_capita")
parameter_means <- parameter_data %>%
  summarise(
    across(all_of(vars_to_standardise),
           ~mean(.x))
  ) %>%
  mutate(measure = "mean") %>%
  rbind(
    parameter_data %>%
      summarise(
        across(all_of(vars_to_standardise),
               ~var(.x))
      ) %>%
      mutate(measure = "var")
  )
parameter_means <- split(parameter_means, parameter_means$measure) %>%
  map(~as.list(.x))

parameter_data <- parameter_data %>%
  mutate(across(
    all_of(vars_to_standardise),
    ~(.x - parameter_means$mean[[cur_column()]])/sqrt(parameter_means$var[[cur_column()]]),
    .names = "{.col}_standardised"
  )) %>%
  select(!all_of(vars_to_standardise))

#make into a model matrix etc
parameter_matrix <- model.matrix(formula(~. -1), parameter_data[, -which(names(parameter_data) == "id")])
#add id back in as rownames
rownames(parameter_matrix) <- parameter_data$id
rm(parameter_data)
```

```{r target data}
target_deaths_averted <- deaths_averted %>%
  ungroup() %>%
  left_join(
    squire::population %>%
      group_by(iso3c) %>%
      summarise(pop = sum(n)),
    by = "iso3c"
  ) %>%
  mutate(
    id = paste(iso3c, type, replicate, sep = "_")
  ) %>%
  group_by(id) %>%
  summarise(
    id = id[1],
    deaths_averted_percentage = sum(deaths_averted)/(pop[1] - sum(deaths)),
    .groups = "drop"
  ) %>%
  mutate(
    deaths_averted_percentage = if_else(
      deaths_averted_percentage < 0,
      0,
      deaths_averted_percentage
    )
  ) %>%
  pull(deaths_averted_percentage, id)
```

```{r fit the candidate models}
set.seed(100100)
#ensure all ids are in the same shuffled order
target_deaths_averted <- sample(target_deaths_averted, length(target_deaths_averted))
parameter_matrix <- parameter_matrix[names(target_deaths_averted), ]
deaths_array <- deaths_array[names(target_deaths_averted),,]
#only do fitting if necessary
if(params$generate_model_fits){
  #use the functional APIr
  
  #input layers
  input_parameters <- layer_input(shape = ncol(parameter_matrix))
  input_deaths <- layer_input(shape = dim(deaths_array)[-1])
  
  #define models
  candidates <- purrr::cross2(list(
      layers = c(100, 100, 100),
      layers = c(10, 50, 100),
      layers = c(100, 50, 10)
    ),
    list(
      activation = "relu",
      activation = "linear",
      activation = "sigmoid"
    )) %>%
    #name lists
    map(function(x){names(x) <- c("layers", "activation"); x}) %>%
    #convert into keras models
    map(~write_model(input_parameters, input_deaths, .x$layers, .x$activation))
  
  ##CV
  n_folds <- 10
  #define the folds
  folds <- get_folds(names(target_deaths_averted), n_folds)
  names(folds) <- names(target_deaths_averted)
  #fit the candidiates seperately on each fold and return the predictions on the
  #left over fold
  candidate_predictions <- map_dfr(seq_len(n_folds), cv_for_super_learner,
    folds, deaths_array, parameter_matrix, target_deaths_averted, candidates)
}
```

```{r train the meta model}
if(params$generate_model_fits){
  #single layer keras aNN
  meta <- keras_model_sequential(input_shape = ncol(candidate_predictions) - 1) %>%
    layer_dense(1, activation = "sigmoid") %>%
    compile(
      loss = "mean_squared_error",
      optimizer = "adam",
      metrics = "mse"
    )
  meta <- fit_keras_model(meta, candidate_predictions %>%
                          select(!y) %>%
                          as.matrix(), candidate_predictions %>% pull(y))
}
```

```{r train the candidates on full data}
if(params$generate_model_fits){
  #train the candidates on the full data set
  candidates_trained <- map(candidates, ~fit_keras_model(.x, list(parameter_matrix, deaths_array), target_deaths_averted))
}
```

```{r combine into the full model}
if(params$generate_model_fits){
  #freeze so they do not train
  candidates_trained <- map(candidates_trained, ~freeze_weights(.x) %>%
                            compile(
                              loss = "mean_squared_error",
                              optimizer = "adam",
                              metrics = "mse"
                            ))
  meta <- freeze_weights(meta)
  inputs <- list(input_parameters, input_deaths)
  final_layer <- layer_concatenate(map(candidates_trained, ~.x(inputs))) %>%
    meta
  super_learner <- keras_model(inputs, final_layer) %>%
    compile(
      loss = "mean_squared_error",
      optimizer = "adam",
      metrics = "mse"
    )
  save_model_tf(super_learner, here::here("analysis/data/derived_data/fit"))
} else {
  super_learner <- load_model_tf(here::here("analysis/data/derived_data/fit"))
}
```

```{r setup data for prediction}
#array for WHO deaths
prediction_deaths_array <- combined_df %>%
  filter(iso3c %in% unique(deaths_averted$iso3c)) %>%
  select(iso3c, date, who_excess_deaths, pop) %>%
  left_join(variant_timings, by = "iso3c") %>%
  left_join(vaccination_start_dates, by = "iso3c") %>%
  mutate(variant = case_when(
      date < start_date_Delta ~ "Wild",
      TRUE ~ "Delta"
    ),
    vaccinations = if_else(
      date < start_date_vaccination,
      "NoVaccine",
      "Vaccine"
  )) %>%
  group_by(iso3c, variant, vaccinations) %>%
  summarise(
    deaths_per_capita = sum(who_excess_deaths[who_excess_deaths > 0])/pop[1],
    .groups = "keep"
  ) %>% #complete vaccinations and deaths
  group_by(iso3c) %>%
  complete(vaccinations = c("Vaccine", "NoVaccine"), variant = c("Wild", "Delta"), fill = list(deaths_per_capita = 0)) %>%
  left_join(death_means, by = c("vaccinations", "variant")) %>%
  mutate(
    deaths_per_capita_standardised = (deaths_per_capita - mean)/sqrt(var)
  ) %>%
  select(!c(mean, var, deaths_per_capita)) %>% 
  convert_df_to_array()

#parameter matrix for prediction
prediction_parameter_matrix <- parameter_matrix
#set epidemic start date to NULL then everything else should be the same
prediction_parameter_matrix[, "start_date_epidemic_standardised"] <- -999
rownames(prediction_parameter_matrix) <- stringr::str_sub(rownames(prediction_parameter_matrix), 1, 3)
prediction_parameter_matrix <- prediction_parameter_matrix[dimnames(prediction_deaths_array)[[1]],]

#calculate new start dates
start_dates_prediction <- combined_df %>%
  filter(iso3c %in% rownames(prediction_parameter_matrix)) %>%
  select(iso3c, who_excess_deaths, date) %>%
  group_by(iso3c) %>%
  arrange(date) %>%
  filter(who_excess_deaths > 0) %>%
  summarise(
    start_date_epidemic = as.numeric(min(date))
  )
iso3cs_prediction <- start_dates_prediction$iso3c
start_dates_prediction <- start_dates_prediction$start_date_epidemic
names(start_dates_prediction) <- iso3cs_prediction
start_dates_prediction <- (start_dates_prediction - parameter_means$mean$start_date_epidemic)/
  sqrt(parameter_means$var$start_date_epidemic)
prediction_parameter_matrix[, "start_date_epidemic_standardised"] <- start_dates_prediction[rownames(prediction_parameter_matrix)]
rm(start_dates_prediction, iso3cs_prediction)
```

```{r predict who deaths averted}
#predict deaths averted
prediction_deaths_averted <- as.numeric(predict(super_learner, list(prediction_parameter_matrix, prediction_deaths_array)))
#now convert to real values
pops <- combined_df %>%
  select(iso3c, pop) %>%
  unique() %>%
  pull(pop, iso3c)
prediction_deaths <- combined_df %>%
  group_by(iso3c) %>%
  summarise(deaths = sum(who_excess_deaths[who_excess_deaths > 0])) %>%
  mutate(deaths = if_else(is.na(deaths), 0, deaths)) %>%
  pull(deaths, iso3c)

who_deaths_averted <- prediction_deaths_averted*(pops[rownames(prediction_parameter_matrix)] - prediction_deaths[rownames(prediction_parameter_matrix)])
names(who_deaths_averted) <- rownames(prediction_parameter_matrix)
```

```{r produce summary tables}
baseline_model_values <- deaths_averted %>% 
  filter(iso3c %in% names(who_deaths_averted)) %>% 
  filter(iso3c != "FSM") %>% 
  #we'll use reported the model values
  group_by(iso3c, type, replicate) %>% 
  summarise(across(c(deaths,deaths_averted),
                   ~sum(.x[.x>0])),
            .groups = "drop_last")
#create per country summary csv
fitted_deaths_averted <- baseline_model_values %>% 
  summarise(across(c(deaths,deaths_averted),
                   ~format_model_est(.x)),
            .groups = "drop") %>% 
  complete(
    iso3c = names(who_deaths_averted), type = c("excess", "reported"), fill = list(
      deaths = "-",
      deaths_averted = "-"
    )
  ) %>% 
  pivot_wider(id_cols = iso3c, names_from = type, values_from = c(deaths, deaths_averted)) %>% 
  #add the estimated data
  mutate(
    deaths_who = format_number(prediction_deaths[iso3c]),
    deaths_averted_who = format_number(who_deaths_averted[iso3c])
  ) %>% 
  transmute(
    `Country` = countrycode::countrycode(iso3c, origin = "iso3c", destination = "country.name"),
    `WHO Excess Deaths` = deaths_who,
    `WHO Excess Deaths\nDeaths Averted (SuperLearner Model)` = deaths_averted_who,
    `Reported Deaths` = deaths_reported, 
    `Reported Deaths\nDeaths Averted (Disease Model)` = deaths_averted_reported,
    `Economist Excess Deaths` = deaths_excess, 
    `Economist Excess Deaths\nDeaths Averted (Disease Model)` = deaths_averted_excess
  )
#create per country summary csv
write_csv(
  fitted_deaths_averted,
  here::here("analysis/data/derived_data/comparison_by_country.csv")
)
#overall summary table:
#large table with all value

summary_df <- baseline_model_values %>% 
  rbind(
    tibble(
      iso3c = unique(baseline_model_values$iso3c),
      deaths = prediction_deaths[unique(baseline_model_values$iso3c)],
      deaths_averted = who_deaths_averted[unique(baseline_model_values$iso3c)],
      type = "who",
      replicate = 1
    )
  ) %>% 
  arrange(iso3c, type, replicate) %>% 
  ungroup() %>% 
  #add the grouping parameters
  mutate(
    who_region = squire.page::get_WHO_region(iso3c),
    income_group = squire.page::get_income_group(iso3c)
  ) %>% #add parameter values
  left_join(
    squire::population %>% 
      group_by(iso3c) %>% 
      summarise(pop = sum(n), .groups = "drop"),
    by = "iso3c"
  ) %>% 
  left_join(vaccinations_total %>% 
              transmute(iso3c = iso3c, doses = first_doses + second_doses), 
            by = "iso3c"
  )


summary_table <- 
  summarise_by(summary_df, NULL, "total") %>% 
  add_row(` ` = "Income Group:") %>% 
  rbind(
    summarise_by(summary_df, "income_group", "total")
  ) %>% 
  add_row(` ` = "WHO Region:") %>% 
  rbind(
    summarise_by(summary_df, "who_region", "total")
  ) %>% #formatting
  mutate(across(everything(),
                ~if_else(is.na(.x), "", .x))) %>% 
  transmute(
    ` ` = ` `,
    `WHO Excess Deaths` = deaths_who,
    `WHO Excess Deaths\nDeaths Averted (SuperLearner Model)` = deaths_averted_who,
    `Reported Deaths` = deaths_reported, 
    `Reported Deaths\nDeaths Averted (Disease Model)` = deaths_averted_reported,
    `Economist Excess Deaths` = deaths_excess, 
    `Economist Excess Deaths\nDeaths Averted (Disease Model)` = deaths_averted_excess
  )
#save
write_csv(summary_table, here::here("analysis/data/derived_data/global_summary.csv"))
```

```{r deaths averted per capita and vaccine}
measure_table <- 
  summarise_by(summary_df, NULL, "total") %>% 
  left_join(
    summarise_by(summary_df, NULL, "capita"),
    by = " "
  ) %>% 
  left_join(
    summarise_by(summary_df, NULL, "vaccine"),
    by = " "
  ) %>% 
  add_row(` ` = "Income Group:") %>% 
  rbind(
    summarise_by(summary_df, "income_group", "total") %>% 
      left_join(
        summarise_by(summary_df, "income_group", "capita"),
        by = " "
      ) %>% 
      left_join(
        summarise_by(summary_df, "income_group", "vaccine"),
        by = " "
      )
  ) %>% 
  add_row(` ` = "WHO Region:") %>% 
  rbind(
    summarise_by(summary_df, "who_region", "total") %>% 
      left_join(
        summarise_by(summary_df, "who_region", "capita"),
        by = " "
      ) %>% 
      left_join(
        summarise_by(summary_df, "who_region", "vaccine"),
        by = " "
      )
  ) %>% 
  relocate(` `) %>% 
  transmute(
    ` ` = ` `,
    `Excess Deaths` = deaths_who,
    `Deaths Averted` = deaths_averted_who,
    `Deaths Averted per 10k people` = deaths_averted_who_per_capita,
    `Deaths Averted per 10k vaccines` = deaths_averted_who_per_vaccine
  ) %>% 
  mutate(across(everything(), ~if_else(is.na(.x), "", .x)))
write_csv(measure_table, here::here("analysis/data/derived_data/global_summary_by_measure.csv"))
```


Ideally we would refit our models to the WHO estimates and compare those results
to the original output. However, the monthly format of the WHO outputs makes them
unsuitable for use in our MCMC method. Instead we will estimate the deaths 
averted from the WHO mortality by exploring the relationship deaths averted and
the Economist values.

We model this relationship by fitting a super learning model, made up of multiple
ANNs implemented in *keras*,
to predict total deaths averted from January 2020 to the 12th of December 2021.
Our chosen features are: total deaths pre vaccination, total deaths once vaccinations
have begun, the start times of the vaccination campaign, and the total numbers
of vaccine doses given by dose. We also include the features to represent the 
varying model inputs used in the vaccine impact paper: dominant vaccine and 
vaccines reported to be used (summarised into types as in the vaccine impact paper)
to represent vaccine efficacy, hospital and ICU bed capacity, income group to
capture effects of mobility, epidemic start date (estimated for WHO estimates as
the first month with positive excess morality), date of introduction for the 
Delta and Omicron variants.

```{r, fig.cap = "Global comparison of estimated WHO deaths averted to deaths averted generated by model fits to reported COVID deaths and process Economist Excess Mortality Estimates."}
knitr::kable(summary_table)
```

```{r, fig.cap = "Global esimates of deaths averted by vaccinations by WB income group or WHO region, total, or per 10k people or vaccinations (in the primary series)."}
knitr::kable(measure_table)
```

# References 

<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->

<div id="refs"></div>

\newpage

### Colophon

<!-- Add something about renv output? -->

This report was generated on `r Sys.time()` using the following computational environment and dependencies: 
Code is available [here](https://github.com/GBarnsley/excesscompare).

```{r colophon, cache = FALSE}
# which R packages and versions?
if ("devtools" %in% installed.packages()) devtools::session_info()
```

The current Git commit details are:

```{r}
# what commit is this file at? 
if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())  
```
